Explanation of run_analysis.R
========================================================

Background
----------

This document explains the generation, structure and content of the tidy_data.txt dataset, derived from the "Human Activity Recognition Using Smartphones Dataset Version 1.0".

The original dataset contains records from experiments involving 30 volunteers aged between 19 and 48 years. Each person performed six activities (WALKING, WALKING_UPSTAIRS, WALKING_DOWNSTAIRS, SITTING, STANDING, LAYING) wearing a smartphone (Samsung Galaxy S II) on the waist. Using its embedded accelerometer and gyroscope, 3-axial linear acceleration and 3-axial angular velocity were capturat a constant rate of 50Hz. The obtained dataset has been randomly partitioned into two sets, where 70% of the volunteers was selected for generating the training data and 30% the test data. 

The sensor signals (accelerometer and gyroscope) were pre-processed by applying noise filters and then sampled in fixed-width sliding windows of 2.56 sec and 50% overlap (128 readings/window). The sensor acceleration signal, which has gravitational and body motion components, was separated using a Butterworth low-pass filter into body acceleration and gravity. The gravitational force is assumed to have only low frequency components, therefore a filter with 0.3 Hz cutoff frequency was used. From each window, a vector of features was obtained by calculating variables from the time and frequency domain.

Each record has the following data:

<ul>
<li>Triaxial acceleration from the accelerometer (total acceleration) and the estimated body acceleration.</li>
<li>Triaxial Angular velocity from the gyroscope.</li>
<li>A 561-feature vector with time and frequency domain variables.</li>
<li>Its activity label.</li>
<li>An identifier of the subject who carried out the experiment.</li>
</ul>

Processing steps
----------------

The tidy_data.txt dataset was generated as follows:

<ol>
<li>Download archive file containing the "Human Activity Recognition Using Smartphones Dataset Version 1.0" into the working directory.</li>
<li>Extract the archived dataset.</li>
<li>Merge the training and the test sets to create one data set.</li>
<li>Extracts only the measurements on the mean and standard deviation for each measurement.</li>
<li>Uses descriptive activity names to name the activities in the data set.</li>
<li>Appropriately labels the data set with descriptive activity names.</li>
<li>Creates a second, independent tidy data set with the average of each variable for each activity and each subject.
</ol>

R code
------

This script requires the following R libraries.

```{r loadRLibs}
library(plyr)
library(reshape)
library(reshape2)
library(gdata)
library(utils)
library(Hmisc)
```

Specify working directory in R. Fetch data from https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip. Extract into working directory. 

```{r loadSourceData}
setwd("C:/Users/ter053/ownCloud/Coursera/Getting and Cleaning Data/project") # Change path as needed
fileURL <- "http://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip"
download.file(fileURL, destfile ="UCI HAR Dataset.zip")
unzip("UCI HAR Dataset.zip", exdir = ".")
```

Next, load data files into R. Begin by reading activity description data into a data frame and assigning column names.


```{r labelData}
act_labels <- read.table("./UCI HAR Dataset/activity_labels.txt", header = F) # read in activity numbers and corresponding descriptions
colnames(act_labels) <- c("id", "description") # add header information (id used for merging later)
```

Read in variable names for training data measurements. Result is a list. 

```{r loadHeaderInfo}
x_labels <- readLines("./UCI HAR Dataset/features.txt") # read in measurement header information
```

Load training measurement, activity and subject data into separate data frames and assign column names.

```{r loadTrainingData}
x_train <- read.table("./UCI HAR Dataset/train/X_train.txt", header=F)  # read in source measurements
colnames(x_train) <- x_labels # add header information
subject_train <- as.data.frame(readLines("./UCI HAR Dataset/train/subject_train.txt"))  # read in subjects
colnames(subject_train) <- "subject" # add header information
activity_train <- read.table("./UCI HAR Dataset/train/y_train.txt", header = F)
colnames(activity_train) <- "id"   # add header information (use for merging later)
```

Combine subject, activity, and measurement training data into a single data frame.  Replace activity number with activity description.

```{r mergeTrainingData}
train <- cbind(subject_train,activity_train,x_train) # combine data frames horizontally
mrg_train <- arrange(join(act_labels,train),id) # replace activity numbers with descriptive strings
```

Load test measurement, activity and subject data into separate data frames and assign column names.

```{r loadTestingData}
x_test <- read.table("./UCI HAR Dataset/test/X_test.txt", header=F)  # read in source measurements
colnames(x_test) <- x_labels # add header information
subject_test <- as.data.frame(readLines("./UCI HAR Dataset/test/subject_test.txt"))  # read in subjects
colnames(subject_test) <- "subject" # add header information
activity_test <- read.table("./UCI HAR Dataset/test/y_test.txt", header = F)
colnames(activity_test) <- "id"   # add header information (use for merging later)
```

Combine subject, activity, and measurement test data into a single data frame.  Replace activity number with activity description.

```{r mergeTestingData}
test <- cbind(subject_test,activity_test,x_test) # combine data frames horizontally
mrg_test <- arrange(join(act_labels,test),id) # replace activity numbers with descriptive strings
```

Cocatenate the training and test data.  Remove unnecessary columns. Subset mean and standard deviation variables. Reorder according to subject.

```{r cocatenateData}
combine <- merge(mrg_train, mrg_test, all.x =T, all.y = T) # create merged data set
combine$id <- NULL # remove "id" as no longer needed
combine <- combine[,c(2,1,3:563)] # reorder columns
extract <- combine[,grep("mean|std|subject|description", names(combine))] # extract subject, activity description, mean and std deviations only
```

Use melt and dcast functions to reshape data - create data summaries sorted by subject and activity description.

```{r reshapeData}
melted <- melt(extract, id=c("subject","description")) # generate skinny data frame
data <- dcast(melted, subject+description ~ variable, mean) # summarise data
data <- data[order(as.numeric(as.character(data$subject))), ]  # sort on subject
```

Remove numerical references from measurement variable names.  Save result as tidy data text file.

```{r tidyData}
names(data) <- gsub("[[:digit:]]", "", names(data)) # remove numbers in measurement headers
names(data) <- gsub("[(\\)\\]", "", names(data)) # remove brackets
write.table(data, file = "./tidy_data.txt", sep = "\t", col.names = T, row.names = F) # generate tidy data set
```


Metadata
--------

**Variables**

```{r varMetadata}
contents.data.frame(data) # describe variables
```

**File structure**

```{r structMetadata}
file <- tempfile() 
write.fwf(x=data, file=file, formatInfo=TRUE) # describe data format
```

End.